# Descriptive statistics {#sec-descriptive}

```{r}
#| include: false

library(tidyverse)
library(here)
library(kableExtra)

library(fontawesome)

library(questionr)



library(readxl)
rses <- read_excel(here("data", "rses.xlsx"))
```


When we have finished this chapter, we should be able to:

::: {.callout-caution icon="false"}
## `r fa("circle-dot", prefer_type = "regular", fill = "red")` Learning objectives

-   Use frequency counts and percentages to describe categorical variables.
-   Display frequency distributions for categorical variables using bar plots.
-   Use statistical measures such as mean, median, standard deviation and interquartile range to describe numerical variables.
-   Generate and assess visualizations like histograms, density plots, and box plots to understand the distribution and spread of numerical data.

:::


## Summarizing categorical data (Frequency Statistics)

The first step in analyzing a categorical variable is to count the occurrences of each label and calculate their frequencies. The resulting collection of these frequencies for all possible categories is known as the **frequency distribution** of the variable. We can also express the frequencies as **proportions** of the total number of observations, which are called relative frequencies or **percentages** (%).

Let's ceate a a frequency table for the `sex` variable:

```{r}
#| echo: false

rses$sex <- factor(rses$sex, labels = c("male", "female"))

freq(rses$sex, cum = T, total = T, valid = F) |> 
  kbl() |> 
  kable_paper("hover")
```

 

The table displays the following:

-   **Absolute frequency (n):** The number of participants in each category (male: 109, female: 149).

-   **Percentage (%):** The proportion of participants in each category relative to the total number of participants (relative frequency) multiplied by $100\%$ (male: $109/258 \times 100\% = 42.2\%$, female: $149/258 \times 100\% = 57.8\%$).

-   **Cumulative percentage (%cum):** The sum of the percentage contributions of all categories up to and including the current one (male: 42.2%, male plus female: 42.2% + 57.8% = 100%).

Note that the percentages in both "$\%$" and "$cum\%$" columns of the table sum up to 100%.











